---
title: "Getting Started with SpNeigh"
author:
  - name: Jinming Cheng
    affiliation:
    - &duk_nus Centre for Quantitative Medicine,
               Duke-NUS Medical School,
               Singapore, 169857, 
               Singapore
    # email: jinming.cheng@outlook.com
    

date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Getting Started with SpNeigh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

output:
  BiocStyle::html_document:
    toc_float: false
bibliography-style: plain
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    eval = TRUE,
    collapse = TRUE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 60,
    fig.height = 4.2,
    fig.width = 5.6
)
```


# Introduction

The **SpNeigh** package provides a flexible and interpretable framework for 
analyzing spatial neighborhoods in high-resolution spatial 
transcriptomics data. 
Designed with support for technologies such as 10x Xenium, Vizgen MERFISH,
and 10x Visium HD, The **SpNeigh** facilitates a range of region-aware and 
distance-aware analyses. Key features include boundary detection, 
spatial weight calculation based on proximity to structural landmarks 
(e.g., region centroid or boundary), spatially-informed differential 
expression testing using spline-based modeling, and visualization of 
local cellular interactions.

By integrating spatial geometry with transcriptomic profiles, 
**SpNeigh** helps uncover biologically meaningful spatial patterns, 
such as boundary-enriched genes, spatial gradients, and 
intercellular interactions, enabling deeper insights into 
tissue organization and cellular communication.


This vignette demonstrates the key functionalities of the **SpNeigh** R package
using a real [10x Xenium Fresh Frozen Mouse Brain Tiny Subset Dataset](https://www.10xgenomics.com/datasets/fresh-frozen-mouse-brain-for-xenium-explorer-demo-1-standard). 

We show how to:

- Identify spatial boundaries and neighboring ring regions
- Explore spatial interaction patterns between clusters
- Perform different expression analysis between two cell populations
- Compute spatial weights from boundary or centroid proximity
- Perform spatial differential expression analysis along spatial gradients
- Perform spatial enrichment analysis along spatial gradients

This workflow highlights the analytical flexibility and biological insights 
enabled by **SpNeigh**.


# Installation

The **SpNeigh** package can be installed from GitHub by using:

```{r, eval=FALSE}
devtools::install_github("jinming-cheng/SpNeigh")
```


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Load packages and data

```{r loadPackage}
library(SpNeigh)
library(ggplot2)
```

Coordinates of mouse brain dataset
```{r}
coords <- readRDS(system.file("extdata", "MouseBrainCoords.rds",
    package = "SpNeigh"
))
head(coords)
```


Ensure the rownames of `coords` are the same with the cell names
```{r}
rownames(coords) <- coords$cell
```


Log normalized expression data generated by `NormalizeData` 
function in *Seurat* package 
```{r}
logNorm_expr <- readRDS(system.file("extdata", "LogNormExpr.rds",
    package = "SpNeigh"
))
class(logNorm_expr)
```


Annotate clusters based on anatomical brain regions
```{r}
new_cell_types <- c(
    "0" = "Meninges",
    "1" = "Cerebral_cortex",
    "2" = "White_matter",
    "3" = "Cerebral_cortex",
    "4" = "Cerebral_cortex",
    "5" = "Thalamus",
    "6" = "Cerebral_cortex",
    "7" = "Hippocampus",
    "8" = "Hippocampus",
    "9" = "Hippocampus",
    "10" = "White_matter",
    "11" = "Cerebral_cortex"
)
```


In this vignette, we focus on cluster-level rather than 
cell type–level analyses.

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Neighborhood analysis for cluster 2 cells

## Detect spatial boundaries

Extract boundaries of cluster 2. Adjust the `eps` and `minPt`s parameters 
to refine the number of subregions detected by the default 
dbscan clustering method.
```{r}
bon_points_c2 <- GetBoundary(
    data = coords,
    one_cluster = 2,
    eps = 120,
    minPts = 10
)
table(bon_points_c2$region_id)
```


Add boundaries of cluster 2 to the spatial plot
```{r PlotWithBoundary_C2, fig.width=6.6, fig.height=4.2}
PlotBoundary(coords, boundary = bon_points_c2)
```

Plot boundary regions
```{r PlotRegion_C2,fig.height = 2.8,  fig.width = 5.6}
bon_polys_c2 <- BuildBoundaryPoly(bon_points_c2)
PlotRegion(bon_polys_c2)
```

## Obtain neighborhood ring regions

Get spatial ring-shaped regions by subtracting the original boundary polygons 
from their corresponding outer buffered polygons. 
The `outer_boundary` is automatically computed when it is not supplied.
```{r}
ring_regions <- GetRingRegion(boundary = bon_points_c2)
```


Plot ring regions
```{r PlotRing_C2,fig.height = 2.8,  fig.width = 5.6}
PlotRegion(ring_regions)
```


## Statistics of cells inside rings

Get cells inside rings for cluster 2. (`GetCellsInside()` takes a few seconds
to run on this example, but may require several minutes or longer 
for larger datasets.)
```{r}
cells_ring <- GetCellsInside(data = coords, boundary = ring_regions)
cells_ring
```


Plot cells inside rings
```{r PlotCellsInsideRing_C2,fig.height = 3.6,  fig.width = 5.6}
PlotCellsInside(cells_ring, point_size = 0.2)
```


Obtain statistics of cells inside rings for cluster 2
```{r}
stats_ring <- StatsCellsInside(cells_ring)
```


Plot proportion of cells in different clusters for each sub region 
using bar plot. 
<!-- The `stat_column` is used to plot "proportion" or "count". -->
```{r PlotStatsBar_Proportion_C2_Ring, fig.height = 4.2,  fig.width = 6.6}
PlotStatsBar(stats_ring, stat_column = "proportion")
```



Plot proportion of cells in different clusters for each sub region 
using donut chart. If `plot_donut = FALSE`, pie chart is used.
```{r PlotStats_Donut_C2_Ring, fig.height = 3.5,  fig.width = 5.6}
PlotStatsPie(stats_ring, plot_donut = TRUE)
```


## Neighborhood interaction of clusters inside rings

Compute neighborhood interaction matrix using K-nearest neighbors for 
cells inside rings
```{r}
interaction_matrix <- ComputeSpatialInteractionMatrix(
    subset(coords, cell %in% cells_ring$cell)
)
interaction_matrix
```


Heatmap of a row-scaled interaction matrix for cells inside rings 
for cluster 2. 
The heatmap reveals that clusters 2, 3, and 4 are spatially adjacent 
to one another. 
```{r Heatmap_InteractionMatrix_C2_Ring, fig.height = 4.2,  fig.width = 5.6}
PlotInteractionMatrix(interaction_matrix)
```

The plot of cells inside rings split by clusters further confirms the 
co-occurrence of clusters 2, 3, 4 in ring region 1.
```{r PlotCellsInside_SplitByCluster_C2_Ring, fig.width=8.4, fig.height=4.2}
PlotCellsInside(cells_ring) +
    facet_wrap(~cluster) +
    Seurat::RotatedAxis() +
    AddBoundaryPoly(ring_regions, linewidth_boundary = 0.1)
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# DE analysis of cluster 2 cells inside and outside boundaries

Most cells in cluster 2 are located in sub region 1. Hence, we focus on the 
DE analysis between cluster 2 cells inside boundary of sub region 1 and 
cells in the outside neighboring ring region 1.


Get cells inside the boundaries of cluster 2
```{r}
cells_inside <- GetCellsInside(data = coords, boundary = bon_points_c2)
cells_inside
```


Obtain cluster 2 cells inside and outside boundary of sub region
```{r}
cells_in <- subset(cells_inside, region_id == 1 & cluster == 2)[["cell"]]
cells_out <- subset(cells_ring, region_id == 1 & cluster == 2)[["cell"]]
```


Perform DE analysis between cells inside boundary and outside boundary 
using the `limma` framewrok (`lmFit` + `eBayes`)
```{r}
tab <- RunLimmaDE(
    exp_mat = logNorm_expr,
    min.pct = 0.25,
    cells_reference = cells_in,
    cells_target = cells_out
)
```


Top DE genes between cells outside boundary and cells inside boundary. 
The DE genes are ordered by the abstract value of `logFC` by default.
```{r}
head(tab[, c("gene", "logFC", "adj.P.Val", "pct.reference", "pct.target")])
```


Set a random seed to make PlotExpression results reproducible
```{r}
set.seed(123)
```


Expression of top DE genes in cluster 2 cells outside boundary. 
The cells are plotted randomly using the given seed when `shuffle = TRUE`. 
If `shuffle = FALSE`, cells with higher expression values 
are plotted last (on the top).
```{r Expression_Slc17a7_c2, fig.width = 8, fig.height = 2.5}
PlotExpression(
    data = coords[colnames(logNorm_expr), ],
    exp_mat = logNorm_expr,
    genes = tab$gene[1:2],
    sub_plot = TRUE,
    one_cluster = 2,
    shuffle = TRUE,
    point_size = 0.1,
    angle_x_label = 45
)
```


Expression of *Slc17a7* for cluster 2 cells inside and outside boundary 
of sub region 1. Cluster 2 cells outside the boundary show much 
higher expression of Slc17a7.
```{r Expression_Slc17a7_InOutBoundary,fig.height = 2.8, fig.width = 5.6}
PlotExpression(
    data = coords[colnames(logNorm_expr), ],
    exp_mat = logNorm_expr,
    genes = "Slc17a7",
    sub_plot = TRUE,
    shuffle = TRUE,
    sub_cells = c(cells_in, cells_out),
    point_size = 0.3
) +
    AddBoundaryPoly(bon_polys_c2[1, ], linewidth_boundary = 0.5)
``` 


Expression of a general marker gene *Sox10* for Oligodendrocytes 
(cluster 2 cells). *Sox10* is expressed in both cluster 2 cells 
inside and outside the boundary. This suggests that cluster 2 cells 
located near the boundary may represent intermediate cell states.
```{r PlotExpression_Sox10,fig.height = 2.8,  fig.width = 5.6}
PlotExpression(
    data = coords[colnames(logNorm_expr), ],
    exp_mat = logNorm_expr,
    sub_cells = c(cells_in, cells_out),
    sub_plot = TRUE,
    genes = "Sox10",
    shuffle = TRUE,
    point_size = 0.3
) +
    AddBoundaryPoly(bon_polys_c2[1, ], linewidth_boundary = 0.5)
```



<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Spatial DE analysis of cells in cluster 0 along spatial weights

## Detect spatial boundaries

Obtain cells in cluster 0
```{r}
cells_c0 <- subset(coords, cluster == 0)[, "cell"]
```


Plot cluster 0 cells without boundary
```{r Plot_NoBoundary_C0, fig.width=8*0.7+1, fig.height=6*0.7}
PlotBoundary(coords[cells_c0, ])
```


Get boundaries of cluster 0
```{r}
bon_points_c0 <- GetBoundary(data = coords, one_cluster = 0)
bon_polys_c0 <- BuildBoundaryPoly(bon_points_c0)
```


Plot boundary edges for cluster 0
```{r PlotBoundary_C0, fig.width=6.6, fig.height=4.2}
PlotEdge(boundary_poly = bon_polys_c0, linewidth_boundary = 0.6)
```


## Compute and plot spatial weights

Compute spatial weights based on the distance to boundaries
```{r}
weights_bon <- ComputeBoundaryWeights(
    data = coords,
    cell_ids = cells_c0,
    boundary = bon_points_c0
)
```


Spatial weights are a named vector, where names correspond to cell IDs
```{r}
weights_bon[1:3]
```


Alternatively, compute spatial weights based on the distance to centroids
```{r}
weights_cen <- ComputeCentroidWeights(data = coords, cell_ids = cells_c0)
```


Plot boundary weights
```{r, fig.width=6.6, fig.height=4.2}
PlotWeights(data = coords, weights = weights_bon, point_size = 0.8) +
    labs(title = "Boundary weights")
```


## Perform spatial differential analysis along boundary weights

Run spatial DE analysis for cluster 0 cells along boundary weights using splines
```{r}
tab_sp <- RunSpatialDE(
    exp_mat = logNorm_expr[, cells_c0],
    spatial_distance = weights_bon,
    cell_ids = cells_c0
)
```


Top DE genes along boundary weights. 
The first spline coefficient (`Z1`) captures the main expression trend 
along the spatial distance. A positive value indicates increasing expression 
with distance, while a negative value indicates decreasing expression.
```{r}
head(tab_sp)
```


Expression of top DE genes along boundary weights
```{r Expression_top_genes_boundary, fig.width = 8, fig.height = 2.5}
PlotExpression(
    data = coords[colnames(logNorm_expr), ],
    exp_mat = logNorm_expr,
    genes = tab_sp$gene[1:2],
    sub_plot = TRUE,
    one_cluster = 0,
    shuffle = TRUE,
    point_size = 0.1,
    angle_x_label = 45
)
```

Plot scaled average expression along boundary weight bins for 
top 10 spatial DE genes. The scaled average expression ranges 
from 0 to 1 by using min-max normalization method.
```{r Expression_along_boundary_weights_c0, fig.width=6.4, fig.height=4}
PlotSpatialExpression(
    exp_mat = logNorm_expr[, cells_c0],
    spatial_distance = weights_bon,
    scale_method = "minmax",
    genes = tab_sp$gene[1:10],
    label_x = "Boundary weights"
)
```


# Spatial enrichment analysis for each gene

Compute spatial enrichment index (SEI) for each gene based on boundary weights. 
The SEI reflects the extent to which gene expression is enriched in 
spatially weighted regions. 
The result is sorted in descending order by `normalized_SEI`.
```{r}
SEI_bon <- ComputeSpatialEnrichmentIndex(
    exp_mat = logNorm_expr[, cells_c0],
    weights = weights_bon
)
head(SEI_bon)
```

Expression of top genes enriched near boundaries
```{r Expression_top_genes_SEI_boundary, fig.width = 8, fig.height = 2.5}
PlotExpression(
    data = coords[colnames(logNorm_expr), ],
    exp_mat = logNorm_expr,
    genes = SEI_bon$gene[1:2],
    sub_plot = TRUE,
    one_cluster = 0,
    shuffle = TRUE,
    point_size = 0.1,
    angle_x_label = 45
)
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

# Session Info

```{r}
sessionInfo()
```

\pagebreak
